Nanocall: An Open Source Basecaller for Oxford Nanopore Sequencing Data

### 2016-03-17 paper-biorxiv/paper.tex
methods intro edits

The main motivation behind Nanocall is to offer an offline alternative to Metrichor, and since the initial signal segmentation step is performed locally, we do not seek to replace it. As such, the input to Nanocall consists of a set of segmented event sequences, stored in ONT-specific \texttt{Fast5} files. Nanocall can also be given a set of state transition probabilities and a set of pore models, but these are optional, and if they are not specified, sensible defaults are used. Nanocall processes each input file separately as follows. It begins by separating the complementary strands in the input event sequence. Next, it computes pore model scaling parameter estimates. Optionally, Nanocall can perform several rounds of training to update the scaling parameters using the Expectation Maximization algorithm, and also to update the state transition parameters using the Baum-Welch algorithm. Finally, Nanocall performs a standard Viterbi decoding on a Hidden Markov Model where the state is the 6-mer in the nanopore, computing the most likely sequence of 6-mers to have generated the given event sequence. The details of the individual steps are given below.

\paragraph{State Transitions.}
The state transitions are prior probabilities of moving from one state to another state. The default state transitions are computed based on 2 parameters: the ``stay'' probability \pstay and the ``skip'' probability \pskip. The former, \pstay, is the probability that 2 consecutive events are emitted from the same context/state. So, this corresponds to a segmentation error where an erroneous event break was introduced. The latter, \pskip, is the probability that 2 consecutive events are emitted from 2 states that differ by more than 1 kmer shift. So, this corresponds to either a segmentation error or a sequencing error (i.e., DNA moved too fast) where 1 or more events were lost. A slight complication is that, by increasing the number of skips, there is always more than 1 way of going from any one state to any other. For example, \texttt{ACGTGT} can be followed by \texttt{GTGTAC} using either 1 or 3 skips. The exact formula for computing state transitions takes this into account this fact:
\begin{align*}
\tau(k_1, k_2) = \quad &
\delta_{k_1 = k_2} \cdot \pstay
\\
& + \delta_{\suffix(k_1, 5) = \prefix(k_2, 5)} \cdot \pstep \cdot \frac{1}{4}
\\
& + \sum_{i=2}^5 \delta_{\suffix(k_1, 6-i) = \prefix(k_2, 6-i)} \cdot \pskipone^{i-1} \cdot \frac{1}{4^i}
\\
& + \sum_{i>5} \pskipone^{i-1} \cdot \frac{1}{4^6}.
\end{align*}
Here: $\delta$ is the standard indicator function; $\mathrm{prefix}(k,i)$ / $\mathrm{suffix}(k,i)$ are the preffix/suffix of $k$ of length $i$; $\pstep := (1 - \pstay - \pskip)$; and $\pskipone = \pskip / (1 + \pskip)$ corresponds to the probability of exactly 1 skip.

>>>>

The main motivation behind Nanocall is to offer an offline alternative to Metrichor, and since the initial signal segmentation step is performed locally, we do not seek to replace it. As such, the input to Nanocall consists of a set of segmented event sequences, stored in ONT-specific \texttt{Fast5} files. Nanocall can also be given a set of state transition probabilities and a set of pore models, but these are optional, and if they are not specified, sensible defaults are used. Nanocall processes each input file separately as follows. It begins by splitting the template and complementary strands into seperate event sequences when a hairpin is found (JTS note: is this accurate?). Next, it estimates the pore model scaling parameters. Optionally, Nanocall can perform several rounds of training to update the scaling parameters using the Expectation Maximization algorithm (REF), and also to update the state transition parameters using the Baum-Welch algorithm (REF). Finally, Nanocall performs Viterbi decoding (REF) of the path through the hidden states, where the state is the 6-mer in the nanopore. The details of the individual steps are given below.

\paragraph{State Transitions.}
The state transitions are prior probabilities of moving from one state to another state. The default state transitions are computed based on 2 parameters: the ``stay'' probability \pstay and the ``skip'' probability \pskip. The former, \pstay, is the probability that 2 consecutive events are emitted from the same context/state. So, this corresponds to a segmentation error where an erroneous event break was introduced. The latter, \pskip, is the probability that 2 consecutive events are emitted from 2 states that differ by more than 1 kmer shift. So, this corresponds to either a segmentation error or a sequencing error (i.e., DNA moved too fast) where 1 or more events were lost. A slight complication is that, by increasing the number of skips, there is always more than 1 way of going from any one state to any other. For example, \texttt{ACGTGT} can be followed by \texttt{GTGTAC} using either 1 or 3 skips. The exact formula for computing state transitions takes this into account this fact:

\begin{align*}
\tau(k_1, k_2) = \quad &
\delta_{k_1 = k_2} \cdot \pstay
\\
& + \delta_{\suffix(k_1, 5) = \prefix(k_2, 5)} \cdot \pstep \cdot \frac{1}{4}
\\
& + \sum_{i=2}^5 \delta_{\suffix(k_1, 6-i) = \prefix(k_2, 6-i)} \cdot \pskipone^{i-1} \cdot \frac{1}{4^i}
\\
& + \sum_{i>5} \pskipone^{i-1} \cdot \frac{1}{4^6}.
\end{align*}
Here: $\delta$ is the standard indicator function; $\mathrm{prefix}(k,i)$ / $\mathrm{suffix}(k,i)$ are the preffix/suffix of $k$ of length $i$; $\pstep := (1 - \pstay - \pskip)$; and $\pskipone = \pskip / (1 + \pskip)$ corresponds to the probability of exactly 1 skip.
